---
title: "Movie Recommendation model for the Movielens dataset"
author: "Vincent"
date: "April 7th 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code is an extract from the Movielens_Report.Rmd file. It contains only the required code to compute the RMSE from the final model selected.

### Loading useful libraries

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
```

### Downloading the data if required

```{r, cache = TRUE}
dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- read.table(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                      col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
```

### Formatting data
```{r}
colnames(movies) <- c("movieId", "title", "genres")
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                           title = as.character(title),
                                           genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")
```

### Splitting the data in a training and a validation test

```{r, message=FALSE}
# Validation set will be 10% of MovieLens data

set.seed(1)
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set

validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set

removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)

```

### Prediction evaluation with RMSE

```{r}
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings - predicted_ratings)^2))
}
```

### final model with regularized movie and user bias

We use the lambda parameter that was previously optimized by a 10-fold cross validation on the edx training set.

```{r}
l <- 4.925

  mu <- mean(edx$rating)
  
  b_i <- edx %>% 
    group_by(movieId) %>%
    summarize(b_i = sum(rating - mu)/(n()+l))
  
  b_u <- edx %>% 
    left_join(b_i, by="movieId") %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - b_i - mu)/(n()+l))

  predicted_ratings <- 
    validation %>% 
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    mutate(pred = mu + b_i + b_u) %>%
    pull(pred)
  
    final_RMSE <- RMSE(predicted_ratings, validation$rating)
    final_RMSE
```

The RMSE for our best model is ***`r final_RMSE`***.
